---
title: "p8105_hw2_qh2283"
output: github_document
date: "2023-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Q1
```{r}
library(tidyverse)

# 1. Clean the data in pols-month.csv
pols <- read.csv("/Users/hoc/Desktop/p8105_hw2_qh2283/fivethirtyeight_datasets/pols-month.csv")

pols_clean <- pols %>%
  # separate 'mon' into 'year', 'month', and 'day'
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  # convert month number to month name
  mutate(month = month(as.integer(month), label = TRUE)) %>%
  # create 'president' variable
  mutate(president = ifelse(!is.na(prez_dem), "dem", "gop")) %>%
  # remove 'prez_dem' and 'prez_gop'
  select(-prez_dem, -prez_gop) %>%
  # remove 'day' variable
  select(-day)
```
```{r}
# 2. Clean the data in snp.csv
snp <- read.csv("/Users/hoc/Desktop/p8105_hw2_qh2283/fivethirtyeight_datasets/snp.csv")
snp_clean <- snp %>%
  # assuming the date column in snp.csv is named 'date'
  separate(date, into = c("year", "month"), sep = "-") %>%
  mutate(month = month(as.integer(month), label = TRUE)) %>%
  arrange(year, month) %>%
  select(year, month, everything())
```

```{r}

# 3. Tidy the unemployment data
unemployment <- read.csv("/Users/hoc/Desktop/p8105_hw2_qh2283/fivethirtyeight_datasets/unemployment.csv")

unemployment_long <- unemployment %>%
  # convert from wide to long format with all columns
  gather(key = "date", value = "unemployment_rate") %>%
  # separate 'date' into 'year' and 'month'
  separate(date, into = c("year", "month"), sep = "-") %>%
  # convert month names to month labels
  mutate(month = month(as.integer(month), label = TRUE))
# Join the datasets
merged_data <- pols_clean %>%
  left_join(snp_clean, by = c("year", "month")) %>%
  left_join(unemployment_long, by = c("year", "month"))
```
Q2
```{r}
library(tidyverse)
library(readxl)

#Import and clean Mr.Trash Wheel data
mr_trash_wheel <- 
  read_excel("202309 Trash Wheel Collection Data.xlsx", sheet = 1, range = "A2:N586") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster))|>
  mutate(homes_powered = (weight_tons * 500) %/% 30)

#Import and clean Professor Trash Wheel data
prof_trash_wheel <- 
  read_excel("202309 Trash Wheel Collection Data.xlsx", sheet = 2, range = "A2:M108") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster))|>
  mutate(homes_powered = (weight_tons * 500) %/% 30)

#Import and clean Gwynnda Trash Wheel data
gwy_trash_wheel <- 
  read_excel("202309 Trash Wheel Collection Data.xlsx", sheet = 4, range = "A2:L157") |>
  janitor::clean_names() |>
  filter(!is.na(dumpster))|>
  mutate(homes_powered = (weight_tons * 500) %/% 30)

#Add a new variable 
mr_trash_wheel <- mr_trash_wheel %>% mutate(wheel_name = "Mr. Trash Wheel")

prof_trash_wheel <- prof_trash_wheel %>% mutate(wheel_name = "Professor Trash Wheel")

gwy_trash_wheel <- gwy_trash_wheel %>% mutate(wheel_name = "Gwynnda Trash Wheel")

#Convert "year" to numeric
mr_trash_wheel$year <- as.numeric(mr_trash_wheel$year)
prof_trash_wheel$year <- as.numeric(prof_trash_wheel$year)
gwy_trash_wheel$year <- as.numeric(gwy_trash_wheel$year)

#Combine all datasets
trash_wheel_collection <- bind_rows(mr_trash_wheel, prof_trash_wheel, gwy_trash_wheel)
str(trash_wheel_collection)


```
Discription\
The `trash_wheel_collection` dataset amalgamates data from three distinct sources: `mr_trash_wheel`, `prof_trash_wheel`, and `gwy_trash_wheel`. It encompasses `r nrow(trash_wheel_collection)` records and `r ncol(trash_wheel_collection)` attributes. Within this dataset, specific columns provide information on collection dates (`date`), total trash weight (`weight_tons`), different trash categories like `plastic_bags`, and a computed variable, `homes_powered`. As per the dataset, Professor Trash Wheel was responsible for collecting `r sum(prof_trash_wheel$weight_tons)` tons of waste. Furthermore, in July 2021, Gwynnda gathered `r sum(gwy_trash_wheel$cigarette_butts[gwy_trash_wheel$month == "July" & gwy_trash_wheel$year == 2021])` cigarette ends.

Q3
```{r}
library(tidyverse)
library(janitor)
library(lubridate)

# Import and clean 
baseline <- 
  read.csv("data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |>
  mutate(
      #convert "sex" and "apoe4" to string
      sex = recode(sex, "1" = "male", "0" = "female"),
      apoe4 = recode(apoe4, "1" = "carrier", "0" = "non-carrier")
  ) |>
  rename("study_id" = "id")
  
# Remove participants who do not meet the criteria 
baseline2 <- filter(baseline, is.na(age_at_onset) | age_at_onset == "." | current_age < age_at_onset)

# Filter the number of participants who developed MCI
mci_developed <- filter(baseline2, !is.na(age_at_onset) & age_at_onset != ".")

w_a_c <- sum(baseline$sex == "female" & baseline$apoe4 == "carrier")
total_women <- sum(baseline$sex == "female")
```

Discription\
After importing the data, we transform the `sex` and `apoe4` variables to render them more interpretably. The initial baseline dataset contains `r nrow(baseline)` entries. Yet, from these, only `r nrow(baseline2)` individuals satisfied the given criteria, as reflected by the record count in the `baseline2` dataset. Additionally, `r nrow(mci_developed)` of the participants were diagnosed with MCI. The mean age of participants at baseline is `r round(mean(baseline2$current_age), digits = 2)`. Within the research, the percentage of female carriers is `r round((w_a_c / total_women)*100, digits = 2)`%.

```{r}
# Import and clean the biomarker values dataset
amyloid <- 
  read.csv("data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names()
amyloid2 <- pivot_longer(amyloid, baseline:time_8, names_to = "time", values_to = "years")

# Combine datasets
combined_mci <- merge(baseline2, amyloid2, by = "study_id")

# Check participants in one of the datasets
only_baseline <- setdiff(baseline2$study_id, amyloid2$study_id)
only_amyloid <- setdiff(amyloid2$study_id, baseline2$study_id)

# Exporting to CSV file
write.csv(combined_mci, "data_mci/combined_mci.csv", row.names = FALSE)
```
In the baseline dataset, there are `r length(only_baseline)` participants. In the amyloid dataset, there are `r length(only_amyloid)` participants. When combined, the new dataset contains `r nrow(combined_mci)` records with `r ncol(combined_mci)` different kinds of data.
